Annotation-Based Empirical Performance Tuning Using Orio

### 2009-02-12 doc/papers/ipdps09/results.tex
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@431 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb

The SpMV operation computes $\forall_{A_{i,j}} \neq 0:y_{i}
\leftarrow y_{i} + A_{i,j} \cdot x_{j}$, where $A$ is a sparse matrix,
and $x$, $y$ are dense vectors. Each element of $A$ is used precisely once
and element reuse is possible for $x$ and $y$. Thus, to optimize SpMV,
one should use compact data structures for $A$ and try to maximize temporal
reuse of $x$ and $y$. One of the most commonly used data structures for
storing a sparse matrix is compressed sparse row (CSR) storage
~\cite{vuduc-thesis},
illustrated in Figure~\ref{fig:spmv}(a). Elements in each row of $A$ are
packed together in a dense array, \texttt{Aval}, and a corresponding array of
integers \texttt{Aind} stores the column indices. The \texttt{Aptr} array
designates where each sparse row begins in \texttt{Aval} and
\texttt{Aind}. An implementation of SpMV using CSR storage is
shown in Figure~\ref{fig:spmv}(b).

PETSc's implementation of SpMV used in this experiment exploits matrix
structure by using \textit{inodes}, which represent rows with identical
nonzero structure.  PETSc detects inodes automatically in arbitrary sparse
matrices. For example, the sparse matrix shown in Figure~\ref{fig:spmv}(a)
contains three inodes. The first inode is of size two because it holds two
adjacent rows (i.e., the first and the second rows) with the same nonzero
structure. The second inode contains three consecutive rows with identical
nonzero structure, and the last inode only has a single row. Knowing the
properties of each inode at runtime enables maximum reuse of vector $x$ since
multiple elements of $x$ can be loaded and used exactly once for all rows in
the same inode structure. To accomplish this, one must use a
register-blocking transformation. To optimize the inode-based SpMV routine,
we incorporated a new transformation module inside Orio that implements
various optimization strategies including register blocking, SIMDization,
memory alignment optimization, loop-control optimization, accumulator
expansion, and thread-level parallelization (with OpenMP). We then used Orio
to automatically select the best version of the optimized inode SpMV. Only
the outer loop that iterates over inodes was parallelized by using OpenMP.

@@

Figure~\ref{fig:ex27-cookie-results} shows the performance results on the
Intel Xeon. The problem size is the number of grid points in the $x$ and $y$
directions, not the size of the sparse input matrix. The matrix dimension is
based on the grid size and the number of field components computed for each grid
point; for example, an $8 \times 8$ problem involves sparse matrices of
dimension 900 with 17,040 nonzero entries. In this experiment, we ran in
three multicore modes: SMP (one MPI process, eight
threads/process), dual (four MPI processes, two threads/process), and virtual
node or VN (eight MPI processes, one thread/process). The code tuned by Orio
consistently outperforms the alternatives. Furthermore, the performance of
the hand-tuned code is almost equivalent to that of the icc-optimized simple
code. Performance differences between the Orio version and the hand-tuned
version become more significant as more threads per process are employed to
facilitate OpenMP parallelization. The SMP and VN results show that
optimizations using loop-level parallelism (through OpenMP) achieve much
better performance than using MPI coarse-grained parallelism.
%Moreover, for the multithread cases, tuning using Orio determined that
%the performance benefit of SIMDization is more significant than that of
%thread-level parallelization only when the problem sizes are small. The
%reason for this is due to the overhead introduced by OpenMP from coordinating
%its threads across processors.
%So, each optimized
%code for the many-thread cases comprises two distinct tuned codes to handle

@@

This section discusses the performance evaluation of the integrated
Pluto-Orio system (discussed in Section~\ref{sec:integration}) on the
multicore Intel Xeon by using a number of application kernels that are
nontrivial to optimize and parallelize. We compare the performance of the
code tuned by Orio with the base code and the Pluto-generated codes. The
Pluto code was obtained by running the base code with
Pluto-0.3.0~\cite{pluto030} using \texttt{--tile} to employ loop tiling for
L1 cache, \texttt{--l2tile} to employ loop tiling for L2 cache, and
\texttt{--unroll} to employ loop unrolling, and for parallel code generation,
an additional \texttt{--parallel} option. We used Pluto's default tile
sizes (L1: 32x32 or 32x32x32; L2: 256x256 or 256x256x256) and default
unroll factors (64 for 1-D unrolling and 8x8 for 2-D
unroll/jamming). All codes were compiled with the Intel C Compiler
using the \texttt{-O3} optimization flag to enable auto-vectorization
and other advanced optimizations, and
\texttt{-parallel} (or \texttt{-openmp} in the case of manual OpenMP parallelization) 
to enable code parallelization.

@@

\subsubsection{2-D Finite-Difference Time-Domain Method for Computational Electromagnetics}  
We consider the two-dimensional Finite Difference Time Domain (FDTD)
algorithm, a popular method for solving the time-dependent Maxwell's
equations in the context of computational electrodynamic problems. As
shown in Figure~\ref{fig:fdtd-2d-code}, the 2-D FDTD method is
implemented as an outer iteration over time containing four
imperfectly nested loops.  The arrays $ex$ and $ey$ denote the
electric field components, and the array $hz$ denotes the magnetic
field.

@@

The performance of the sequential 2-D FDTD code for $tmax=500$ and $nx=ny$ is
shown in Figure~\ref{fig:fdtd-2d-cookie-seq}. The base code optimized by icc
alone performs better than Pluto for small problem sizes since all input
arrays fit in the L2 cache (insufficient computation to offset Pluto's tiling
overhead). As the array sizes increase, the lack of data reuse impairs the
base code's performance, whereas the Pluto performance remains about the
same.
%We used Orio to generate two code variants: one for small problem sizes
%and another for large problem sizes. 
When the input arrays are small, Orio discovers that applying
Pluto's polyhedral transformations is not beneficial, and therefore
it employs only its syntactic transformations on the original FDTD
code. For large problem sizes, Orio exploits some of the Pluto's code
transformations and enhances them further with its syntactic
optimizations, resulting in performance consistently and significantly
higher than both the base and Pluto codes (up to 86\% over Pluto).

@@

Figure~\ref{fig:fdtd-2d-cookie-par} shows the multicore performance
obtained for $tmax=500$ and $nx=ny=2000$. The results indicate that
whereas icc is unable to auto-parallelize the code, Pluto detects the
existence of pipelined parallelism and then successfully parallelizes
the code. Locality-exploiting optimizations by Orio additionally
improve the Pluto performance by up to 78\%. Moreover, we observe
that because of memory contention between the two quad-core Intel
processors, the speedup of both the Pluto and Orio codes slightly
decreases when the number of cores used is greater than four.

@@

\subsubsection{3-D Gauss-Seidel Successive Over-Relaxation Method}  
Figure~\ref{fig:seidel-code} shows the 3-D Gauss-Seidel computation, 
which is sometimes referred to as
\emph{successive displacement method}, indicating the dependence of the 
iterations on the ordering. If the ordering is changed, the components
of the new iterations will change as well.
%
%\begin{figure}[ht]
\begin{figure}%[ht]
\begin{center}
\begin{minipage}{3in} 
\scriptsize
\begin{verbatim} 
for (t=0; t<=T-1; t++) 
  for (i=1; i<=N-2; i++) 
    for (j=1; j<=N-2; j++) 
      A[i][j] = (A[i-1][j-1] + A[i-1][j] 
          + A[i-1][j+1] + A[i][j-1] + A[i][j] 
          + A[i][j+1] + A[i+1][j-1] + A[i+1][j] 
          + A[i+1][j+1]) / 9.0; 
\end{verbatim} 
\end{minipage} 
\end{center}
\caption{3-D Gauss-Seidel code} 
\label{fig:seidel-code} 
\end{figure}
%

@@

LU factorization or decomposition is a numerical method for the
solution of linear systems of equations; a simple implementation is
shown in Figure~\ref{fig:lu-code}.
%
\begin{figure}%[ht]
\begin{center}
\begin{minipage}{2.8in} 
\scriptsize
\begin{verbatim} 
for (k=0; k<=N-1; k++) { 
  for (j=k+1; j<=N-1; j++) 
    A[k][j] /= A[k][k]; 
  for(i=k+1; i<=N-1; i++) 
    for (j=k+1; j<=N-1; j++) 
      A[i][j] -= A[i][k]*A[k][j]; 
} 
\end{verbatim} 
\end{minipage} 
\end{center}
\caption{LU Decomposition code.} 
\label{fig:lu-code} 
\end{figure}
%

>>>>

The SpMV operation computes $\forall_{A_{i,j}} \neq 0:y_{i}
\leftarrow y_{i} + A_{i,j} \cdot x_{j}$, where $A$ is a sparse matrix,
and $x$, $y$ are dense vectors. Each element of $A$ is used precisely once,
and element reuse is possible only for $x$ and $y$. Thus, to optimize SpMV,
one should use compact data structures for $A$ and try to maximize temporal
reuse of $x$ and $y$. One of the most commonly used data structures for
storing a sparse matrix is compressed sparse row (CSR) storage
~\cite{vuduc-thesis},
illustrated in Figure~\ref{fig:spmv}(a). Elements in each row of $A$ are
packed together in a dense array, \texttt{Aval}, and a corresponding array of
integers \texttt{Aind} stores the column indices. The \texttt{Aptr} array
designates where each sparse row begins in \texttt{Aval} and
\texttt{Aind}. An implementation of SpMV using CSR storage is
shown in Figure~\ref{fig:spmv}(b).

PETSc's implementation of SpMV used in this experiment exploits matrix
structure by using \textit{inodes}, which represent rows with identical
nonzero structure.  PETSc detects inodes automatically in arbitrary sparse
matrices. For example, the sparse matrix shown in Figure~\ref{fig:spmv}(a)
contains three inodes. The first inode is of size two because it holds two
adjacent rows (i.e., the first and the second rows) with the same nonzero
structure. The second inode contains three consecutive rows with identical
nonzero structure, and the last inode has only a single row. Knowing the
properties of each inode at runtime enables maximum reuse of vector $x$ since
multiple elements of $x$ can be loaded and used exactly once for all rows in
the same inode structure. To accomplish this, one must use a
register-blocking transformation. To optimize the inode-based SpMV routine,
we incorporated a new transformation module inside Orio that implements
various optimization strategies including register blocking, SIMDization,
memory alignment optimization, loop-control optimization, accumulator
expansion, and thread-level parallelization (with OpenMP). We then used Orio
to automatically select the best version of the optimized inode SpMV. Only
the outer loop that iterates over inodes was parallelized by using OpenMP.

@@

Figure~\ref{fig:ex27-cookie-results} shows the performance results on the
Intel Xeon. The problem size is the number of grid points in the $x$ and $y$
directions, not the size of the sparse input matrix. The matrix dimension is
based on the grid size and the number of field components computed for each grid
point; for example, an $8 \times 8$ problem involves sparse matrices of
dimension 900 with 17,040 nonzero entries. In this experiment, we ran in
three multicore modes: SMP (one MPI process, eight
threads/process), dual (four MPI processes, two threads/process), and virtual
node or VN (eight MPI processes, one thread/process). The code tuned by Orio
consistently outperforms the others. Furthermore, the performance of
the hand-tuned code is almost equivalent to that of the icc-optimized simple
code. Performance differences between the Orio version and the hand-tuned
version become more significant as more threads per process are employed to
facilitate OpenMP parallelization. The SMP and VN results show that
optimizations using loop-level parallelism (through OpenMP) achieve much
better performance than using MPI coarse-grained parallelism.
%Moreover, for the multithread cases, tuning using Orio determined that
%the performance benefit of SIMDization is more significant than that of
%thread-level parallelization only when the problem sizes are small. The
%reason for this is due to the overhead introduced by OpenMP from coordinating
%its threads across processors.
%So, each optimized
%code for the many-thread cases comprises two distinct tuned codes to handle

@@

This section discusses the performance evaluation of the integrated
Pluto-Orio system (discussed in Section~\ref{sec:integration}) on the
multicore Intel Xeon by using a number of application kernels that are
nontrivial to optimize and parallelize. We compare the performance of the
code tuned by Orio with the base code and the Pluto-generated codes. The
Pluto code was obtained by running the base code with
Pluto-0.3.0~\cite{pluto030} using \texttt{--tile} to employ loop tiling for
L1 cache, \texttt{--l2tile} to employ loop tiling for L2 cache, and
\texttt{--unroll} to employ loop unrolling, and for parallel code generation,
an additional \texttt{--parallel} option. We used Pluto's default tile
sizes (L1: 32x32 or 32x32x32; L2: 256x256 or 256x256x256) and default
unroll factors (64 for 1-D unrolling and 8x8 for 2-D
unroll/jamming). All codes were compiled with the Intel C compiler
using the \texttt{-O3} optimization flag to enable auto-vectorization
and other advanced optimizations, and
\texttt{-parallel} (or \texttt{-openmp} in the case of manual OpenMP parallelization) 
to enable code parallelization.

@@

\subsubsection{2-D Finite-Difference Time-Domain Method for Computational Electromagnetics}  
We consider the two-dimensional finite difference time domain (FDTD)
algorithm, a popular method for solving the time-dependent Maxwell's
equations in the context of computational electrodynamic problems. As
shown in Figure~\ref{fig:fdtd-2d-code}, the 2-D FDTD method is
implemented as an outer iteration over time containing four
imperfectly nested loops.  The arrays $ex$ and $ey$ denote the
electric field components, and the array $hz$ denotes the magnetic
field.

@@

The performance of the sequential 2-D FDTD code for $tmax=500$ and $nx=ny$ is
shown in Figure~\ref{fig:fdtd-2d-cookie-seq}. The base code optimized by icc
alone performs better than Pluto for small problem sizes since all input
arrays fit in the L2 cache (insufficient computation to offset Pluto's tiling
overhead). As the array sizes increase, the lack of data reuse impairs the
base code's performance, whereas the Pluto performance remains about the
same.
%We used Orio to generate two code variants: one for small problem sizes
%and another for large problem sizes. 
When the input arrays are small, Orio discovers that applying
Pluto's polyhedral transformations is not beneficial, and therefore
it employs only its syntactic transformations on the original FDTD
code. For large problem sizes, Orio exploits some of the Pluto's code
transformations and enhances these further with its syntactic
optimizations, resulting in performance consistently and significantly
higher than both the base and Pluto codes (up to 86\% over Pluto).

@@

Figure~\ref{fig:fdtd-2d-cookie-par} shows the multicore performance
obtained for $tmax=500$ and $nx=ny=2000$. The results indicate that
whereas icc is unable to autoparallelize the code, Pluto detects the
existence of pipelined parallelism and then successfully parallelizes
the code. Locality-exploiting optimizations by Orio additionally
improve the Pluto performance by up to 78\%. Moreover, we observe
that because of memory contention between the two quad-core Intel
processors, the speedup of both the Pluto and Orio codes slightly
decreases when the number of cores used is greater than four.

@@

\subsubsection{3-D Gauss-Seidel Successive Overrelaxation Method}  
Figure~\ref{fig:seidel-code} shows the 3-D Gauss-Seidel computation, 
which is sometimes referred to as
\emph{successive displacement method}, indicating the dependence of the 
iterations on the ordering. If the ordering is changed, the components
of the new iterations will change as well.
%
%\begin{figure}[ht]
\begin{figure}%[ht]
\begin{center}
\begin{minipage}{3in} 
\scriptsize
\begin{verbatim} 
for (t=0; t<=T-1; t++) 
  for (i=1; i<=N-2; i++) 
    for (j=1; j<=N-2; j++) 
      A[i][j] = (A[i-1][j-1] + A[i-1][j] 
          + A[i-1][j+1] + A[i][j-1] + A[i][j] 
          + A[i][j+1] + A[i+1][j-1] + A[i+1][j] 
          + A[i+1][j+1]) / 9.0; 
\end{verbatim} 
\end{minipage} 
\end{center}
\caption{3-D Gauss-Seidel code} 
\label{fig:seidel-code} 
\end{figure}
%

@@

LU factorization or decomposition is a numerical method for the
solution of linear systems of equations; a simple implementation is
shown in Figure~\ref{fig:lu-code}.
%
\begin{figure}%[ht]
\begin{center}
\begin{minipage}{2.8in} 
\scriptsize
\begin{verbatim} 
for (k=0; k<=N-1; k++) { 
  for (j=k+1; j<=N-1; j++) 
    A[k][j] /= A[k][k]; 
  for(i=k+1; i<=N-1; i++) 
    for (j=k+1; j<=N-1; j++) 
      A[i][j] -= A[i][k]*A[k][j]; 
} 
\end{verbatim} 
\end{minipage} 
\end{center}
\caption{LU decomposition code.} 
\label{fig:lu-code} 
\end{figure}
%
